---
title: "cism: The CISM R package"
author: "Joe Brew"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.width = 6,
                      fig.height = 4)
```

`cism`, the R package of the Centro de Investigação em Saude de Manhiça, is a collaborative effort to improve the way that CISM researchers interact with and utilize data. 

## Objectives

The objectives of the R CISM package are simple:

1. Make retrieving data more simple.

2. Make data analysis more reproducible.

3. Make interaction with data more pleasurable and beautiful.

## Set up

Before using the `cism` package, you'll need to (1) install and (2) configure credentials.

### 1. Install

To use the `cism` package, you'll first need to install it. With a good internet connection, run the following from within R.

```r
if(!require(devtools)) install.packages("devtools")
install_github('joebrew/cism')
```

Once you've installed the package, you can use its functionality in any R script by simply including the following line:

```r
library(cism)
```

### 2. Configure credentials

Accessing CISM databases requires permissions. All the functionality in the `cism` package can be used by calling functions with a user's credentials. However, both for the sake of ease as well as to avoid putting any confidential information into R code, it is highly recommended to create a `credentials` folder with a `credentials.yaml` file. The contents of this file should look something like this:

```
dbname: openhds
host: sap.manhica.net
port: 3306 # Use 4706 from outside of the CISM
user: your_user_name
password: your_pass_word
```

In the above, consider replacing `dbname` with the database you most often use. Replace `port` with 4706 if you work outside of the CISM. And replace the `user` and `password` fields with your name and password. If you use git, or any other form of version control, ensure that the `credentials.yaml` file is ignored by any repository viewable by others.

The `credentials/credentials.yaml` will be automatically detected by the `cism` package if in the same directory as the working directory, or if it is in a higher directory. Therefore, one should either create a `credentials/credentials.yaml` for each project, or place one `credentials/credentials.yaml` into a higher-level folder (for example "My Documents" for Windows users, or "/home/" for Linux users), under which all code resides.

## Basic use

### Retrieving data from CISM databases

With the `cism` package, getting data is simple. Assuming you've got a `credentials/credentials.yaml` in or above your working directory, you can simply point the `get_data` function at the database and table of your choice:

```r
residency <- get_data(tab = 'residency', dbname = 'openhds')
```

The above will return the entire "residency" table from the "openhds" database.

If you're a more advanced user and want to filter, select or join directly in SQL before collecting the results into memory, you can write directly in SQL:

```r
small_residency <- get_data(query = "SELECT uuid, startDate FROM residency limit 10", dbname = 'openhds')
```

### Visualizing data

The `cism` package offers several tools for the visualization of data. 

#### cism_plot

`cism_plot` is a simple wrapper for plotting a variable of any type, or 2 numeric variables. Its use is relatively straightforward.


We'll start by creating some fake data.


```{r}
library(sp)
library(cism)
# Create a fake dataset of people
people <- data.frame(id = 1:1000,
                     sex = sample(c('Male', 'Female', 'Unknown'),
                                  size = 1000,
                                  replace = TRUE),
                     latitude = rnorm(mean = coordinates(man3)[,2],
                                      sd = 0.1,
                                      n = 1000),
                     longitude = rnorm(mean = coordinates(man3)[,1],
                                       sd = 0.1,
                                       n = 1000),
                     age = sample(seq(22, 
                                      26,
                                      length = 10000), 
                                  size = 1000, replace = TRUE),
                     money = sample(seq(0,
                                        100000,
                                        length = 1000000), size = 1000, replace = TRUE))
```

`cism_plot` intuitively handles the visualization of counts of categorical data:

```{r, message = FALSE}
cism_plot(x = people$sex)
```

It also handles numeric variables.

```{r}
cism_plot(x = people$age)
```

If explicitly told to, it will treat a numeric variable as a categorical.

```{r}
cism_plot(x = round(people$age), type = 'factor')
```

Finally, if given two (numeric) variables, CISM plot will generate a simple x-y chart.

```{r}
cism_plot(x = people$age,
          y = people$money)
```

#### cism_map

Like `cism_plot`, `cism_map` strives to generate useful and aesthetically pleasing maps with the least amount of input necessary. In the below examples, we'll use the `people` dataset we generated above.

At its most simple level, `cism_map` can show the location of longitude and latitude.

```{r, message = FALSE}
cism_map(lng = people$longitude,
         lat = people$latitude)
```

By providing a "fortified SpatialPolygonsDataFrame" (of which several come with the `cism` package), the map will get adminsitrative outlines for the relevant areas.

```{r, message = FALSE}
cism_map(lng = people$longitude,
         lat = people$latitude,
         fspdf = man3_fortified)
```

If we want to see our points colored by a categorical variable, we simply indicate the variable in our call to `cism_map`.

```{r, message = FALSE}
cism_map(lng = people$longitude,
         lat = people$latitude,
         fspdf = man3_fortified,
         x = people$sex)
```

By the same token, points can also be colored by a continuous variable.

```{r, message = FALSE}
cism_map(lng = people$longitude,
         lat = people$latitude,
         fspdf = man3_fortified,
         x = people$money)
```

As with other ggplot objects, plots generated by `cism_map` or `cism_plot` can be given further modifications through additional calls to ggplot2 functions. For example:

```{r, message = FALSE}
cism_map(lng = people$longitude,
         lat = people$latitude,
         fspdf = man2_fortified,
         x = people$sex) +
  ggtitle('Spatial distribution by sex')
```


#### cism_map_interactive

`cism_map_interactive` is very similar to `cism_map`. Rather than build static maps, it builds interactive satellite-based maps.

To use, you'll need a dataset with at least longitude and latitude. Here we'll again use our `people` dataset:

```{r}
cism_map_interactive(lng = people$longitude, 
                     lat = people$latitude)
```

We can add a SpatialPolygonsDataFrame, including those included in the cism package, with the `spdf` arugment.

```{r}
cism_map_interactive(lng = people$longitude, 
                     lat = people$latitude, 
                     spdf = man3)
```

Our points can be simple representations of location (above), or we can color them by a categorical variable.

```{r}
cism_map_interactive(lng = people$longitude,
                     lat = people$latitude,
                     spdf = man2,
                     x = people$sex)
```

Rather than a categorical variable (above), we can also color points by a continous variable.

```{r}
cism_map_interactive(lng = people$longitude,
                     lat = people$latitude,
                     spdf = man2,
                     x = people$age)
```

### Built-in datasets

The `cism` package also includes several built-in datasets. 

Not finished.

## Advanced use

Not finished